White Wine Quality Analysis by Divya Chandramouli
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(GGally)
library(scales)
library(gridExtra)
library(RColorBrewer)
library(knitr)
```


```{r}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```


Load the data and print the names of variables in the dataset

```{r echo=FALSE, Load_the_Data}
# Load the Data
wq <- read.csv('wineQualityWhites.csv')
names(wq)
```

# Introduction to the dataset

This data set contains data on white variants of the Portuguese "Vinho Verde" 
wine. There are 11 variables quantifying the chemical properties of each wine. 
At least 3 wine experts rated the quality of each wine, providing a rating between 
0 (very bad) and 10 (very excellent).This dataset is publicly available for 
research.

Citation:  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

The inputs include objective tests (e.g. PH values) and the output is based on 
sensory data (median of at least 3 evaluations made by wine experts).

# Description of Attributes

1 - fixed acidity: most acids involved with wine or fixed or nonvolatile 
  (do not evaporate readily)

2 - volatile acidity: the amount of acetic acid in wine, which at too high of 
  levels can lead to an unpleasant, vinegar taste

3 - citric acid: found in small quantities, citric acid can add 'freshness' and
  flavor to wines

4 - residual sugar: the amount of sugar remaining after fermentation stops, 
   it's rare to find wines with less than 1 gram/liter and wines with greater 
   than 45 grams/liter are considered sweet

5 - chlorides: the amount of salt in the wine

6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between 
  molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
  growth and the oxidation of wine

7 - total sulfur dioxide: amount of free and bound forms of S02; in low 
 concentrations, SO2 is mostly undetectable in wine, but at free SO2 
 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

8 - density: the density of water is close to that of water depending on the 
 percent alcohol and sugar content

9 - pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
 to 14 (very basic); most wines are between 3-4 on the pH scale

10 - sulphates: a wine additive which can contribute to sulfur dioxide gas 
 (S02) levels, wich acts as an antimicrobial and antioxidant

11 - alcohol: the percent alcohol content of the wine

   Output variable (based on sensory data): 
   
12 - quality (score between 0 and 10)

# Research Question

This project investigates which chemical properties influence the quality of 
white wines

# Univariate Plots Section

In this section, we will perform some preliminary exploration of the dataset- 

Running some summaries of the data and creating univariate plots to
understand the structure of the individual variables in the dataset. 

### Dimensions of the dataset

```{r}
dim(wq)
```
We have 4898 rows and 13 variables in the dataset - 11 physicochemical 
properties, 1 quality rating and 1 index/serial number


### Let's examine the structure of the dataframe, the variables and the type of
data they hold


```{r echo=FALSE, Univariate_Plots}

str(wq)

```

### Summary of the dataset

```{r}

summary(wq)

```

### Check for missing or null values in the dataset

```{r}

any(is.na(wq))

```

We do not any have missing values in the dataset
 
 
### Are there duplicate records in the dataset? 
 
```{r}

sum(duplicated(wq))

```
 
 No, the dataset does not have duplicate records
 
### Individual variable exploration
 
#### Fixed acidity 
 
```{r}
table(wq$fixed.acidity)

```
 
 
 
```{r}

ggplot(aes(x= fixed.acidity), data = wq)+geom_histogram(binwidth = 0.05)+ 
  scale_x_continuous(breaks=seq(3,15),limits = c(3,15))

```
 
 The distribution of fixed acidity looks like the bell curve and appears to be 
 normally distributed. The range of fixed acidity is from 3.8 to 14.2 from the 
 summary table. In the plot we can clearly see that the fixed acidity values for
 the bulk of wines lies between 4 and 10 with outliers on the right. 6.8 is the 
 most commonly occurring value of fixed acidity 
 
 
#### Volatile acidity 
  
```{r}
table(wq$volatile.acidity)
```
  
 
```{r}

ggplot(aes(x= volatile.acidity), data = wq)+geom_histogram(binwidth = 0.001)

```
 
 The distribution of volatile acidity looks skewed to the right. 
 Most common value of volatile acidity is 0.28. Bulk of the 
 wines have volatile acidity value under 0.6.
 
 Since the distribution is skewed to the right, let us try a log10 transform 
 
```{r}
ggplot(aes(x= volatile.acidity), data = wq)+geom_histogram(binwidth = 0.001)+scale_x_log10()
```
 
 Now the distribution looks closer to the bell curve of normal distribution
 
 
 #### Citric acid
 
```{r}
table(wq$citric.acid)
```
 
 
```{r}
ggplot(aes(x= citric.acid), data = wq)+geom_histogram(binwidth=0.002)
```
 
 Bulk of the citric acid data looks normally distributed. There is an outlier 
 with the value 1.66
 
 #### Residual sugar
 
```{r}
table(wq$residual.sugar)

```
 
 
```{r}
ggplot(aes(x= residual.sugar), data = wq)+geom_histogram(binwidth=0.1)
```
 
 The distribution of residual sugar looks clearly skewed to the right with the 
 maxvalue of 65.8 being an outlier
 
 Let's try a log10 transform on the data
 
```{r}

ggplot(aes(x= residual.sugar), data = wq)+geom_histogram(binwidth=0.1)+scale_x_log10()
```
 
 Now the distribution looks bimodal with peaks at about 1 and 9 values of residual sugar
 
 If we trim the top 1% of residual sugar data, let's see how the distribution looks
 
```{r}
 ggplot(aes(x= residual.sugar), data = wq)+geom_histogram(binwidth=0.1)+
  xlim(0, quantile(wq$residual.sugar, probs = 0.99))
```
 
 The distribution is still skewed to the right after trimming the top 1% of data
 
 #### Chlorides
 
```{r}

table(wq$chlorides)
```
 
 
```{r}
ggplot(aes(x= chlorides), data = wq)+geom_histogram(binwidth=0.0005)
```
 
 The distribution of chloride values looks skewed to the right with a thin and 
 long tail on the right
 
 
 Let us trim the top 3% of chloride values and see how our data looks
 
```{r}

ggplot(aes(x= chlorides), data = wq)+geom_histogram(binwidth=0.0005)+
  xlim(0, quantile(wq$chlorides, probs=0.97))
```
 
 
 Now the data looks a little closer to normal distribution. But it might not be
 a good idea to lose 3% of the data
 
 
 #### Free Sulphur dioxide
 
```{r}
table(wq$free.sulfur.dioxide)
```
 
```{r}
ggplot(aes(x= free.sulfur.dioxide), data = wq)+geom_histogram(binwidth =1)
```
 The distribution of free sulphurdioxide looks skewed to the right with an 
 outlier point at  289
 
 #### Total Sulphur dioxide
```{r}
table(wq$total.sulfur.dioxide)
```
 
```{r}
ggplot(aes(x= total.sulfur.dioxide), data = wq)+geom_histogram(binwidth=1)
```
 The bulk of total sulphur dioxide data looks normally distributed with a 
 few outliers on the right
 
 #### Density
 
```{r}
ggplot(aes(x= density), data = wq)+geom_histogram(binwidth=0.0001)
```
The distribution of density looks almost normally distributed 


```{r}
summary(wq$density)
```

Few wines seem to have a density greater than 1 

 #### pH

```{r}
table(wq$pH)
```


```{r}
ggplot(aes(x= pH), data = wq)+geom_histogram(binwidth=0.001)
```

pH values appear to be normally distributed. From domain knowledge, we can 
expect pH / acidity to be key factors influencing the quality of wine 

 #### Sulphates
 
```{r}
table(wq$sulphates)
```
 
```{r}
ggplot(aes(x= sulphates), data = wq)+geom_histogram(binwidth=0.01)
```
 
 The distribution of sulphates appears skewed to the right
 
 Let us transform the sulphates axis on a log10 scale
 
 
```{r}
ggplot(aes(x= sulphates), data = wq)+geom_histogram(binwidth=0.01)+
  scale_x_log10()
```
The distribution of sulphates appears closer to normal distribution with a log10
transformation
 
 #### Alcohol
```{r}
table(wq$alcohol)
```
 
```{r}
ggplot(aes(x= alcohol), data = wq)+geom_histogram(binwidth=0.01)
```
 
 The histogram for alcohol level does not look normally distributed.
 
 let us try a log10 transform on alcohol level data
 
```{r}
ggplot(aes(x= alcohol), data = wq)+geom_histogram()+
  scale_x_log10()
```
 
Even after the log10 transformation, alcohol level data does not look normally
distributed

Let us try a sqrt transform on the alcohol level data

```{r}
ggplot(aes(x= alcohol), data = wq)+geom_histogram()+
  scale_x_sqrt()
```

The sqrt transformation also doesn't seem to make the alcohol level data 
normally distributed

From common knowledge, alcohol levels of wine could be a key influencer of wine 
quality

 #### Quality
 
```{r}
table(wq$quality)
```
 
 
```{r}
ggplot(aes(x= quality), data = wq)+geom_histogram()
```
 
 From the quality rating histogram, we can see that the bulk of wines have been 
 rated 6, and only 5 wines have received a quality rating of 9, which is the best
 
 The histogram consists of discrete bars at 3,4,5,6,7,8 and 9.  
 
 
# Univariate Analysis


### What is the structure of your dataset?

The dataset contains 4898 rows/observations with 13 variables/columns
eleven of the columns represent physicochemical properties of wines and hold 
numerical values, one of the columns is the index/ID of each row which is an
integer and  one of the columns holds the quality rating for the wine which 
is also an integer

### What is/are the main feature(s) of interest in your dataset?

The goal of this project is to investigate the influence of the physicochemical 
properties on the quality of wine. Therefore, quality rating is one of the main 
features of interest along with some other properties. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Since quality rating is from sensory data, residual sugar and pH could affect the
sweetness/taste of wine and alcohol content could affect the experience of wine
consumption. But the investigation will look at all the properties in the 
dataset


### Did you create any new variables from existing variables in the dataset?

No, from just univariate exploration, it could not be ascertained if new variables 
need to be created.It will be looked at if further exploration gives rise to 
the need for creating more variables

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The data was assessed for missing or duplicate values. It appeared fairly tidy 
and prepared for analysis. Some distributions were heavily skewed, so I used
log transformation of data and tried trimming a portion of the data to look at 
the distribution 

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

Let us look at a scatterplot matrix to explore relationships between variables. 
Let us leave 'quality' (it is a categorical ordinal variable)
and 'X' (which is the index/count) out.  

```{r echo=FALSE, Bivariate_Plots}
wq_1 = wq[,c(2:12)]
ggpairs(wq_1)
```

Let's look at the correlation coefficicents for all variables in the dataset. 
It does not make sense to include the 'X' variable which is the index, when we
want to look at correlations. Since quality is a categorical ordinal variable, 
the Spearman correlation coefficient should be calculated.
So, I am using the dataframe excluding X and quality



The correlation coefficients for all variables in the dataset rounded to one
decimal place:

```{r}
corr_coeff=round(cor(wq_1),1)
corr_coeff
```

From these correlation coefficients, the values that are greater than |0.3| are 
interpreted to be more meaningful as they signify a moderate/strong relationship


```{r}
corr_coeff[abs(corr_coeff) > 0.3]
```
Below are the variables and correlation coefficients with absolute values
that are > 0.3

```{r}
zcor <- as.data.frame(as.table(corr_coeff))
m_zcor <- subset(zcor,abs(Freq)>=0.3)
kable(m_zcor)
```

We see some moderate and some strong relationships from the table above:

- Fixed acidity and pH have a moderate negative correlation, which can be reasoned 
out as pH value decreases with increasing acidity

- Residual sugar has a strong positive correlation with density which makes sense 
as more sugar makes the wine denser

- Free and total sulphur dioxide have a strong positive relationship which is 
expected, because total sulphur dioxide is the sum of free and bound sulphurdioxide

- Alcohol and density have a strong negative relationship suggesting that higher 
alcohol content results in lesser wine density . Alcohol is less dense than water, 
so this relationship seems to make sense

Let us visualize the relationship between residual sugar and density since it is
strong

```{r}
ggplot(aes(x=density, y=residual.sugar), data=wq)+geom_point()
```


Let's limit the axis, reduce overplotting and look at the plot

```{r}
ggplot(aes(x=density, y=residual.sugar), data=wq)+geom_point(alpha=1/10)+scale_x_continuous(limits=c(0.98, 1.015))
```



### Spearman's correlation coefficients for quality and other variables
Spearman's correlation is generally used when ordinal variables are invloved. 
In this case we want to correlate continuous variables with an ordinal variable.
For the purpose of this course, I have used Spearman's coefficient but there 
could be more complex but more suited methods to accomplish this.
The following are the Spearman correlation coefficients for all variables with quality:

```{r}
quality_cor <-cor(wq[, c(2:13)],
    method = 'spearman')[ ,12]

quality_cor<-as.data.frame(quality_cor)
print(quality_cor)
```

From the above values we find that 
 - Quality and Chlorides are moderately negatively correlated
 - Quality and density are moderately negatively correlated
 - Quality and alcohol are positively correlated 
 
Let's explore the relationships between Quality and a few variables in more detail

Summary statistics for wine density
```{r}
summary(wq$density)
```


The following is the summary statistics of density grouped by quality
```{r}
by(wq$density, wq$quality,summary)
```

From the statistics above we see that higher quality wines seem to have lesser density

Let us create a boxplot of density and quality

```{r}
ggplot(aes(x=factor(quality), y=density), data=wq)+geom_boxplot(fill='pink')+stat_summary(funy = mean, color='blue',geom='point', size=2)
```

The blue dot marks the mean density of each quality level of wine



Boxplot of alcohol level and quality

```{r}
ggplot(aes(x=factor(quality), y=alcohol), data=wq)+geom_boxplot(fill='green')+stat_summary(funy = mean, color='black',geom='point', size=2)
```

```{r}
by(wq$alcohol, wq$quality,summary)

```



```{r}
by(wq$residual.sugar,wq$quality,summary)
```



```{r}
ggplot(aes(x=factor(quality), y=residual.sugar), data=wq)+geom_boxplot(fill='purple')+stat_summary(funy = mean, color='yellow',geom='point', size=2)
```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}


```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!