---
output:
  html_document: default
  pdf_document: default
---
White Wine Quality Analysis - Divya Chandramouli
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(GGally)
library(scales)
library(gridExtra)
library(RColorBrewer)
library(knitr)
library(psych)
library(memisc)
```

### Loading data

Load the data and print the names of variables in the dataset

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wq <- read.csv('wineQualityWhites.csv')
names(wq)
```

### Introduction to the dataset

This dataset contains data on white variants of the Portuguese "Vinho Verde" 
wine. There are 11 variables quantifying the chemical properties of each wine. 
At least 3 wine experts rated the quality of each wine, providing a rating between 
0 (very bad) and 10 (very excellent).This dataset is publicly available for 
research.

Citation:  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

The inputs include objective tests (e.g. PH values) and the output is based on 
sensory data (median of at least 3 evaluations made by wine experts).

### Description of Attributes

1. Fixed acidity: most acids involved with wine or fixed or nonvolatile 
  (do not evaporate readily)

2. Volatile acidity: the amount of acetic acid in wine, which at too high of 
   levels can lead to an unpleasant, vinegar taste

3. Citric acid: found in small quantities, citric acid can add 'freshness' and
   flavor to wines

4. Residual sugar: the amount of sugar remaining after fermentation stops, 
   it's rare to find wines with less than 1 gram/liter and wines with greater 
   than 45 grams/liter are considered sweet

5. Chlorides: the amount of salt in the wine

6. Free sulfur dioxide: the free form of SO2 exists in equilibrium between 
   molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
   growth and the oxidation of wine

7. Total sulfur dioxide: amount of free and bound forms of S02; in low 
   concentrations, SO2 is mostly undetectable in wine, but at free SO2 
   concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

8. Density: the density of water is close to that of water depending on the 
   percent alcohol and sugar content

9. pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
   to 14 (very basic); most wines are between 3-4 on the pH scale

10. Sulphates: a wine additive which can contribute to sulfur dioxide gas 
   (S02) levels, wich acts as an antimicrobial and antioxidant

11. Alcohol: the percent alcohol content of the wine

   
    Output variable (based on sensory data): 
   
12. Quality (score between 0 and 10)

### Research Question

This project investigates which chemical properties influence the quality of 
white wines

### Understanding the dataset

In this section, we will perform some preliminary exploration of the dataset- 
Running some summaries of the data and creating univariate plots to
understand the structure of the individual variables in the dataset. 

#### Structure of the dataset

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(wq)
```
We have 4898 rows and 13 variables in the dataset - 11 physicochemical 
properties, 1 quality rating and 1 index/serial number


Let's examine the structure of the dataframe, the variables and the type of
data they hold


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}

str(wq)

```

#### Summary of the dataset

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(wq)

```

#### Check for missing or null values in the dataset

```{r echo=FALSE, message=FALSE, warning=FALSE}

any(is.na(wq))

```

We do not any have missing values in the dataset
 
 
#### Are there duplicate records in the dataset? 
 
```{r echo=FALSE, message=FALSE, warning=FALSE}

sum(duplicated(wq))

```
 
 No, the dataset does not have duplicate records
 
### Univariate exploration
 
#### Fixed acidity 
 
 
```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x= fixed.acidity), data = wq)+geom_histogram(binwidth = 0.05)+ 
  scale_x_continuous(breaks=seq(3,15),limits = c(3,15))

```
 
 The distribution of fixed acidity looks like the bell curve and appears to be 
 normally distributed. The range of fixed acidity is from 3.8 to 14.2 from the 
 summary table. In the plot we can clearly see that the fixed acidity values for
 the bulk of wines lies between 4 and 10 with outliers on the right. 6.8 is the 
 most commonly occurring value of fixed acidity 
 
 
#### Volatile acidity 

 
```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x= volatile.acidity), data = wq)+geom_histogram(binwidth = 0.001)

```
 
 The distribution of volatile acidity looks skewed to the right. 
 Most common value of volatile acidity is 0.28. Bulk of the 
 wines have volatile acidity value under 0.6.
 
 Since the distribution is skewed to the right, let us try a log10 transform 
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= volatile.acidity), data = wq)+geom_histogram(binwidth = 0.001)+scale_x_log10()
```
 
 Now the distribution looks closer to the bell curve of normal distribution
 
 
#### Citric acid
 
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= citric.acid), data = wq)+geom_histogram(binwidth=0.002)
```
 
 Bulk of the citric acid data looks normally distributed. There is an outlier 
 with the value 1.66
 
#### Residual sugar

 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar), data = wq)+geom_histogram(binwidth=0.1)
```
 
 The distribution of residual sugar looks clearly skewed to the right with the 
 maxvalue of 65.8 being an outlier
 
 Let's try a log10 transform on the data
 
```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x= residual.sugar), data = wq)+geom_histogram(binwidth=0.1)+scale_x_log10()
```
 
 Now the distribution looks bimodal with peaks at about 1 and 9 values of residual sugar
 
 If we trim the top 1% of residual sugar data, let's see how the distribution looks
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
 ggplot(aes(x= residual.sugar), data = wq)+geom_histogram(binwidth=0.1)+
  xlim(0, quantile(wq$residual.sugar, probs = 0.99))
```
 
 The distribution is still skewed to the right after trimming the top 1% of data
 
#### Chlorides
 
````{r echo=FALSE ,message=FALSE, warning=FALSE}
ggplot(aes(x= chlorides), data = wq)+geom_histogram(binwidth=0.0005)

````
 


The distribution of chloride values looks skewed to the right with a thin and 
long tail on the right
 
 
 Let us trim the top 3% of chloride values and see how our data looks
 
```{r echo=FALSE ,message=FALSE, warning=FALSE}

ggplot(aes(x= chlorides), data = wq)+geom_histogram(binwidth=0.0005)+
  xlim(0, quantile(wq$chlorides, probs=0.97))
```
 
 
 Now the data looks a little closer to normal distribution. But it might not be
 a good idea to lose 3% of the data
 
 
#### Free Sulphur dioxide
 
 
```{r echo=FALSE ,message=FALSE, warning=FALSE}
ggplot(aes(x= free.sulfur.dioxide), data = wq)+geom_histogram(binwidth =1)
```
 

 The distribution of free sulphurdioxide looks skewed to the right with an 
 outlier point at  289
 
#### Total Sulphur dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= total.sulfur.dioxide), data = wq)+geom_histogram(binwidth=1)
```
 
 
 The bulk of total sulphur dioxide data looks normally distributed with a 
 few outliers on the right
 
#### Density
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= density), data = wq)+geom_histogram(binwidth=0.0001)
```


The distribution of density looks almost normally distributed 
Few wines seem to have a density greater than 1 - most probably they should be 
outliers

#### pH

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= pH), data = wq)+geom_histogram(binwidth=0.001)
```

pH values appear to be normally distributed. From domain knowledge, we can 
expect pH / acidity to be key factors influencing the quality of wine 


#### Sulphates
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= sulphates), data = wq)+geom_histogram(binwidth=0.01)
```
 
 The distribution of sulphates appears skewed to the right
 
 Let us transform the sulphates axis on a log10 scale
 
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= sulphates), data = wq)+geom_histogram(binwidth=0.01)+
  scale_x_log10()
```

The distribution of sulphates appears closer to normal distribution with a log10
transformation
 
#### Alcohol
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol), data = wq)+geom_histogram(binwidth=0.01)
```
 
 The histogram for alcohol level does not look normally distributed.
 
 Let us try a log10 transform on alcohol level data
 
```{r echo=FALSE ,message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol), data = wq)+geom_histogram()+
  scale_x_log10()
```
 
Even after the log10 transformation, alcohol level data does not look normally
distributed

Let us try a sqrt transform on the alcohol level data

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol), data = wq)+geom_histogram()+
  scale_x_sqrt()
```

The sqrt transformation also doesn't seem to make the alcohol level data 
normally distributed

From common knowledge, alcohol levels of wine could be a key influencer of wine 
quality

#### Quality
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality), data = wq)+ geom_histogram()
```
 
 From the quality rating histogram, we can see that the bulk of wines have been 
 rated 6, and only 5 wines have received a quality rating of 9, which is the best
 
 The histogram consists of discrete bars at 3,4,5,6,7,8 and 9.  
 
 
### Univariate Analysis


#### What is the structure of your dataset?

The dataset contains 4898 rows/observations with 13 variables/columns
eleven of the columns represent physicochemical properties of wines and hold 
numerical values, one of the columns is the index/ID of each row which is an
integer and  one of the columns holds the quality rating for the wine which 
is also an integer

#### What is/are the main feature(s) of interest in your dataset?

The goal of this project is to investigate the influence of the physicochemical 
properties on the quality of wine. Therefore, quality rating is one of the main 
features of interest along with some other properties. 

#### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Since quality rating is from sensory data, residual sugar and pH could affect the
sweetness/taste of wine and alcohol content could affect the experience of wine
consumption. But the investigation will look at all the properties in the 
dataset


#### Did you create any new variables from existing variables in the dataset?

No, from just univariate exploration, it could not be ascertained if new variables 
need to be created.It will be looked at if further exploration gives rise to 
the need for creating more variables

#### Of the features you investigated, were there any unusual distributions? 
#### Did you perform any operations on the data to tidy, adjust, or change the form 
#### of the data? If so, why did you do this?

The data was assessed for missing or duplicate values. It appeared fairly tidy 
and prepared for analysis. Some distributions were heavily skewed, so I used
log transformation of data and tried trimming a portion of the data to look at 
the distribution 

### Bivariate Exploration

Let us look at a scatterplot matrix to explore relationships between variables. 
Let us leave 'X' (which is the index/count) out.  


```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
wq_1 = wq[,c(2:12)]
ggpairs(wq[,c(2:13)])+
theme_minimal()
```




We get a fair idea of the relationships from the scatterplot matrix but it would
be useful to go into the Pearson correlation coefficients for the variables excluding 
quality
It does not make sense to include the 'X' variable which is the index, when we
want to look at correlations. 
Since quality is a categorical ordinal variable, the Spearman correlation
coefficient should be calculated.
So I am using the dataframe excluding X and quality to look at 
Pearson coefficients


Let us generate a correlation matrix to get a visual idea about the strength of the relationships

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggcorr(wq_1, label=TRUE,label_color='black')
```

Let's look closely at the correlation coefficicents from the above visual 

The values that are greater than |0.3| are interpreted to be more meaningful as 
they signify a moderate/strong relationship

We see some moderate and some strong relationships from the visual above:

- Fixed acidity and pH have a moderate negative correlation, which can be reasoned 
out as pH value decreases with increasing acidity

- Residual sugar has a strong positive correlation with density which makes sense 
as more sugar makes the wine denser

- Free and total sulphur dioxide have a strong positive relationship which is 
expected, because total sulphur dioxide is the sum of free and bound sulphurdioxide

- Alcohol and density have a strong negative relationship suggesting that higher 
alcohol content results in lesser wine density . Alcohol is less dense than water, 
so this relationship seems to make sense

Let's plot the moderate correlations 

Fixed acidity vs pH


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=fixed.acidity, y=pH), data=wq)+geom_point(alpha=1/10)+geom_smooth(method='lm')
  
```

Free Sulphur dioxide vs Total sulphur dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide), data=wq)+geom_point(alpha=1/10)+geom_smooth(method='lm')+scale_x_continuous(limits=c(0,200))
```

Let us visualize the strong relationships:

Residual sugar vs density 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y=residual.sugar), data=wq)+geom_point()
```


Let's limit the axis to exclude outliers, reduce overplotting and look at the 
plot with a linear trendline

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y=residual.sugar), data=wq)+geom_point(alpha=1/40)+
  scale_x_continuous(limits=c(0.98, 1.015))+
  scale_y_continuous(limits=c(0,40))+geom_smooth(method='lm')
```

Let us visualize the relationship between alcohol and density

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol, y=density), data=wq)+geom_point()
```

Let's reduce overplotting and look at the plot again with a linear trendline,
excluding two outliers
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol, y=density), data=subset(wq, wq$density<1.01))+
  geom_point(alpha=1/20,position='jitter')+
  geom_smooth(method = 'lm')
  
```


#### Spearman's correlation coefficients for quality with other variables
Spearman's correlation is generally used when ordinal variables are invloved. 
In this case we want to correlate continuous variables with an ordinal variable.
For the purpose of this course, I have used Spearman's coefficient but there 
could be more complex but more suited methods to accomplish this.
The following are the Spearman correlation coefficients for all variables with 
quality:

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(wq[, c(2:13)],
    method = 'spearman')[ ,12]
```

From the above values we find that 
 - Quality and Chlorides are moderately negatively correlated
 - Quality and density are moderately negatively correlated
 - Quality and alcohol are positively correlated 
 
### Creating new variables

Discrete Quality Variable

It would be helpful to create a discrete quality variable and view the summary
```{r echo=FALSE, message=FALSE, warning=FALSE}
wq$quality.discrete=factor(wq$quality)
summary(wq$quality.discrete)
```
Quality Bucket Variable

Let's also create a quality bucket variable

Summary of quality variable
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$quality)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
wq$quality.bucket <- cut(wq$quality, breaks=c(0,4,7, 10), 
                      labels=c('low', 'moderate', 'high'))
```

Summary of quality buckets:
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$quality.bucket)
```

Residual sugar bucket variable

Summary statistics for residual sugar
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$residual.sugar)
```
Let's create a residual.sugar.bucket variable and view the summary

```{r echo=FALSE, message=FALSE, warning=FALSE}
wq$residual.sugar.bucket <- cut(wq$residual.sugar, breaks = c(-Inf, 2,7.5,Inf), 
                                labels= c('Low', 'Medium','High'))
summary(wq$residual.sugar.bucket)
```

Alcohol Bucket Variable

Let's create an alcohol.bucket variable
```{r echo=FALSE, message=FALSE, warning=FALSE}
wq$alcohol.bucket <- cut(wq$alcohol, breaks=c(0,9.5, 12, 15), 
                      labels=c('low', 'moderate', 'high'))
```

Summary Statistics for the alcohol bucket variable:
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$alcohol.bucket)
```


Density Bucket Variable

Summary of density
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$density)
```

Let's create a density.bucket variable
```{r echo=FALSE, message=FALSE, warning=FALSE}
wq$density.bucket <- cut(wq$density, breaks=c(-Inf,0.992, 0.9955, Inf), 
                      labels=c('low', 'medium', 'high'))
```

Summary Statistics for the alcohol bucket variable:
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$density.bucket)
```


#### Quality vs Chloride
Let's plot discrete quality ratings and chloride values to visualize the relationship

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality.discrete, y=chlorides), data=wq)+geom_boxplot(aes(color=quality.discrete))+stat_summary(fun.y=mean,geom='point')
```

The relationship is not too  clear from the box plot. However, the mean Chloride 
values for the high quality wines are slightly lesser than the lower quality wines


#### Quality vs Density

Let's look at the boxplot of density vs quality, excluding some density outliers
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality.discrete, y=density), data=wq)+geom_boxplot(aes(color=quality.discrete))+stat_summary(fun.y=mean,geom='point')+coord_cartesian(ylim=c(0.985,1.01))
```


Summary statistics for wine density

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$density)
```

Results for density statistics by quality level:
```{r echo=FALSE, message=FALSE, warning=FALSE}
with(wq,by(density,quality,summary))
```




Summary statistics for density by quality.bucket:
```{r echo=FALSE, message=FALSE, warning=FALSE}
with(wq,by(density,quality.bucket,summary))
```


Summary statistics for density bucket by quality bucket:

```{r echo=FALSE, message=FALSE, warning=FALSE}
with(wq,by(density.bucket,quality.bucket,summary))
```

About 90% of high quality wines fall in the low and medium density buckets 
(60% - low density and 30% - medium density bucket)
Only about 10% of high quality wines have high density.
In general, we see that the mean  density for high quality wines is lesser than 
that of low quality wines


#### Quality vs Alcohol

Let's look at the boxplot of alcohol vs quality
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality.discrete, y=alcohol), data=wq)+geom_boxplot(aes(color=quality.discrete))+stat_summary(fun.y=mean,geom='point')
```

It's interesting to see a clear trend of increasing alcohol content as
the quality of wine gets better. 

Summary Statistics for Alcohol bucket variable

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$alcohol)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wq$quality.bucket)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(wq,quality >7)$alcohol > 10.51)
```

More than 80% of high quality wines have alcohol content greater than the mean 
alcohol content in wines

Summary Statistics for Alcohol bucket by Quality bucket
```{r echo=FALSE, message=FALSE, warning=FALSE}
with(wq, by(alcohol.bucket,quality.bucket,summary))
```
Only 5% of low quality wines have high alcohol content, while about 
50% of high quality wines have high alcohol content

### Bivariate Analysis

#### Talk about some of the relationships you observed in this part of the investigation.
#### How did the feature(s) of interest vary with other features in the dataset?

From the bivariate exploration, it was seen that quality, which is the
main feature of interest did not strongly correlate with any of the variables. 
Alcohol level, density and chlorides seemed to correlate moderately with 
quality. 

#### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
Though quality did not directly correlate strongly with the variables, alcohol level 
vs density  and  residual sugar vs density are strong relationships.
Fixed acidity acidity correlated negatively with pH which makes sense from knowledge
of general chemistry

High quality wines seem to have high alcohol content and low density

#### What was the strongest relationship you found?

- Density vs residual sugar (0.8) 
- Density vs alcohol (-0.8)

### Multivariate Exploration
Let's pick the strongest relationships we saw in bivariate analysis and add a 
third variable to the plots

I'm curious about residual sugar since it is a parameter that can affect taste 
distinctly. So in the plot of density vs residual sugar, let us see how adding 
information on the quality bucket helps us understand the relationship
The outliers for density have been excluded

#### Density vs residual sugar by quality bucket

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(x=density, y=residual.sugar), data=subset(wq,wq$density < 1.01))+geom_point(aes(color=quality.bucket))+
scale_color_brewer(type='seq')
```

The gradient in color denoting quality is not very clear. There is overplotting 
and the moderate quality wines are dominant

Let us subset the data to look at only low and high quality wines to see if we 
can get clearer insight


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y=residual.sugar), data= subset(wq, wq$density < 1.01 & wq$quality.bucket != 'moderate'))+
geom_point(aes(color=quality.bucket))

```
For a fixed value of density, the higher quality wines have more residual sugar 
than the lower quality wines which is interesting. 

Now let us color the density vs residual sugar plot by alcohol bucket


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y=residual.sugar), data=subset(wq,wq$density < 1.01))+geom_point(aes(color=alcohol.bucket))+
scale_color_brewer(type='seq')
```
Here the gradient in color denoting alcohol bucket looks very interesting
For a fixed density, wines with higher residual sugar have greater alcohol content 
At higher levels of residual sugar, the density is higher and alcohol content is 
lower


Let us facet the density vs residual sugar plot colored by alcohol
using quality.discrete

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=density, y=residual.sugar), data=subset(wq,wq$density < 1.01))+geom_point(aes(color=alcohol.bucket))+
scale_color_brewer(type='seq')+facet_wrap(~quality.discrete)
```
The plots with lower quality ratings are dominated by wines with lower alcohol
As the quality rating increases, wines have higher alcohol and lower density


#### Density vs alcohol level, by quality bucket

Let us add quality bucket tto our density vs alcohol plot excluding density 
outliers
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol, y=density), data=subset(wq, wq$density<1.01))+
  geom_point(aes(color=quality.bucket),position='jitter')+
  scale_color_brewer(type="seq")
```

Again we see that the moderate quality wines are dominant
Let us subset the data to include only low and high quality wines


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol, y=density), data=subset(wq, wq$density<1.01 & wq$quality.bucket != 'moderate'))+
  geom_point(aes(color=quality.bucket))
```

This plot shows us that wines with higher quality have higher alcohol content 
and lower density. At lower alcohol levels, the variation in density is more

Let us color the density vs alcohol plot by residual sugar 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol, y=density), data=subset(wq, wq$density<1.01))+
  geom_point(aes(color=residual.sugar.bucket),position='jitter')+
  scale_color_brewer(type="seq")
```

We can seethat wines with higher residual sugar have higher density for the same
alcohol content

Now let's facet this graph by quality

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol, y=density), data=subset(wq, wq$density<1.01))+
  geom_point(aes(color=residual.sugar.bucket),position='jitter')+
  scale_color_brewer(type="seq")+facet_wrap(~quality.discrete)
```


The variation in density decreases as quality gets better. the plots with medium quality ratings
are dominated by wines with high residual sugar


Since alcohol vs quality was the strongest relationship we saw among all
the relationships with quality, let's explore that by coloring it by residual sugar

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality.discrete, y=alcohol), data=wq)+geom_point(aes(color=residual.sugar.bucket))
```

The buckets of residual sugar seem to be distributed among the levels of quality, 
with a slight domination on the higher side of quality. Not much can be ascertained 
about the influence of residual sugar on quality.

As quality increases the alcohol content increases and the
variation in alcohol content decreases 

#### Quality Prediction Model

Let's try to build a mdel to predict quality given the values for a few properties.
Since quality is not strongly correlated with any of the features the accuracy of the model cannot be expected to be high

Let's try using alcohol as the first feature since the correlation with alcohol was highest for quality, The other features I'm choosing are: density, chlorides and residual sugar
```{r echo=FALSE, message=FALSE, warning=FALSE}
m1 <- lm(I(quality)~I(alcohol), data=wq)
m2 <- update(m1, ~ . + density)
m3 <- update(m2, ~ . + chlorides)
m4 <- update(m3, ~ . + residual.sugar)

mtable(m1, m2, m3, m4)

```
The R square values are very poor which implies we will have to adopt a better 
approach to building a predictive model for quality. Using more relevant features 
and complex methods may help build a better predictive model for quality

### Multivariate Analysis

#### Talk about some of the relationships you observed in this part of the investigation.
#### Were there features that strengthened each other in terms of looking at your feature(s) of interest?


Quality was the main feature of interest. Density and alcohol levels seemed to influence quality and the relationships were explored more in detail by introducing bucket variables for alcohol, quality, density and residual sugar. But since none of the relationships were strong enough to build a predictive model for quality, the R squared values for the linear  model are poor 


#### Were there any interesting or surprising interactions between features?


It was interesting to observe the variation in density / alcohol levels as the quality gets better. Faceting plots by quality and coloring them with a feature helped to understand
and explain some trends

#### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

I tried to create a linear predictive model for quality using alcohol, density, chlorides and residual sugar. Since the strength of relationships with quality was not great, the model does very poorly. More data with relevant features and more complex methods may help to build a better model.

------

### Final Plots and Summary

#### Plot One 
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
p1 <- ggplot(aes(x=quality.discrete, y=alcohol), data=wq)+geom_boxplot(aes(color=quality.discrete))+stat_summary(fun.y=mean,geom='point')+
  xlab('Quality Rating')+ylab('Alcohol level (% by volume)')+
  ggtitle('Alcohol levels by Quality rating')

p2<- ggplot(aes(x=quality.discrete, y=density), data=wq)+geom_boxplot(aes(color=quality.discrete))+stat_summary(fun.y=mean,geom='point')+
  xlab('Quality Rating')+ylab('Density (g/cubic cm)')+
  ggtitle('Density by Quality rating')

grid.arrange(p1,p2,ncol=2)
```

#### Description One
This plot clearly illustrates that higher quality wines have higher alcohol content.
The black dots denote the mean alcohol content for every quality level. The increasing trend in alcohol content is very pronounced from rating 5 - 9. The strongest correlation observed with quality was that of alcohol content with a Spearman's coefficient of 0.44

#### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x=alcohol, y=density), data=subset(wq, wq$density<1.01 & wq$quality.bucket != 'moderate'))+
  geom_point(aes(color=quality.bucket))+ggtitle('Alcohol level vs Density by Quality bucket')+
xlab('Alcohol level (% by volume)')+
  ylab('Density (g/cubic cm)')



```

#### Description Two
This graph is colored by quality bucket and shows only the low and high quality wines
It is interesting to see that higher alcohol content corresponds to lesser density (alcohol is less denser than water). From Plot one and two, we can infer that higher quality wines  have higher alcohol and lesser density 

#### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x=alcohol, y=density), data=subset(wq, wq$density<1.01))+
  geom_point(aes(color=residual.sugar.bucket),position='jitter')+
  scale_color_brewer(type="seq")+
  xlab('Alcohol level (% by volume)')+
  ylab('Density (g/cubic cm)')+
  ggtitle('Residual sugar buckets in density vs alcohol plot')

```


#### Description Three
The variance in density with decreasing alcohol content is accounted for by the 
residual sugar levels in wine. 
Density, alcohol and residual sugar were part of strong correlations and this 
plot summarizes their relationship. When this plot was faceted by quality it could 
further help in picturing the variation of these parameters with quality.

------

### Reflection

Univariate exploration helped get an idea about the individual distributions of 
data and it was interesting to apply transformations  and see their effect.  
I had expected residual sugar to be a direct influencer of taste and therefore
quality since the quality rating is sensory data. It turned out that 
residual sugar correlated with density strongly and explained the variance in 
density over alcohol and quality levels. 

The interesting part of the exploration was investigating the strong relationships 
between density and residual sugar, density and alcohol and connecting them to 
quality ratings using colored and faceted plots.

The linear predictive model for quality was built using alcohol, density, chlorides 
and residual sugar. Since the strength of relationships with quality is not great, 
the model does very poorly. More data with relevant features and more complex 
methods may help to build a better model


### Resources

https://www.rdocumentation.org/packages/GGally/versions/1.3.2/topics/ggpairs

https://www.rdocumentation.org/packages/memisc/versions/0.99.14.9/topics/mtable

https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt


